---
title: "Course Project 1"
author: "Crystal Kizer"
date: "10/21/2020"
output: html_document
---
Loading and preprocessing the data:

```{r}
#Loading Data to read file
setwd("C:/Users/AF03941/Documents/Coursera")
activity<-read.csv("activity.csv")
```
Process/transform the data into a format suitable for your analysis
```{r}
#process data. Review activity file to understand the dataset
dim(activity) 
names(activity) 
head(activity) 
str(activity) 


#formats date 
activity$date <- as.Date(activity$date, "%Y-%m-%d")
#remove NA from data file
head(activity[which(!is.na(activity$steps)), ]) 
```
##What is mean total number of steps taken per day?


```{r}
# Calculate the total number of steps taken per day
total_steps <- aggregate(steps ~ date, activity, sum )
names(total_steps)[names(total_steps)=="steps"] <- "total_steps"
 total_steps 

```

```{r}
# Make a histogram of the total number of steps taken each day
sumSteps <- aggregate(activity$steps ~ activity$date, FUN=sum) 
colnames(sumSteps)<- c("Date", "Steps") 
 hist(sumSteps$Steps, breaks=10, xlab="Steps", main = "Total Steps per day") 
```

Calculate and report the mean and median of the total number of steps taken per day
```{r}
#MEAN Steps per day:
as.integer(mean(sumSteps$Steps)) 
#MEDIAN Steps per day:
as.integer(median(sumSteps$Steps)) 

```
What is the average daily activity pattern?

```{r} 
#Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

 #calculate mean steps by 5 min interval
x <- aggregate(steps ~ interval, activity, mean, na.rm=TRUE) 

 # plot time series chart and label accordingly
 plot(x, type="l", xlab="5 Min Interval", ylab="AVG Steps Taken") 
```   

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r} 
# get the interval which has the most steps
x$interval[x$steps==max(x$steps)]
``` 
Imputing missing value:
Calculate and report the total number of missing values in the dataset 
```{r} 
sum(is.na(activity$steps))
``` 
Devise a strategy for filling in all of the missing values in the dataset. Create a new dataset that is equal to the original dataset but with the missing data filled in

```{r} 
# create the new dataset to fill in values for NA 
 Fillin <- activity 
# substitute NA with zeros in dataset
 Fillin$steps[is.na(Fillin$steps)] <- 0 
# make sure there are no NAs in dataset
 sum(is.na(Fillin$steps)) 
# validate the dataset is equal to the original dataset but with the missing data filled in as 0s
sum(!is.na(Fillin$steps)) 
```  

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 
 

```{r} 
TotalSteps2 <- aggregate(Fillin$steps ~ Fillin$date, FUN=sum) 
colnames(TotalSteps2)<- c("Date", "Steps") 

  #Create histogram of total number of steps
 hist(TotalSteps2$Steps, breaks=10, xlab="Steps", main = "Total Steps per day") 
#report the mean and median total number of steps taken per day
#Mean steps per day
as.integer(mean(TotalSteps2$Steps)) 
#Median steps per day
as.integer(median(TotalSteps2$Steps)) 

#Do these values differ from the estimates from the first part of the assignment? 
#Yes, filling in the missing values to 0 skewed the data closer to 0. We can see there are more 0 values in our historgram.  
#What is the impact of imputing missing data on the estimates of the total daily number of steps?
#Decreased both mean and median

```
Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
#create weekday and weekend variable 
activity$dayType <- sapply(activity$date, function(x) {
  if(weekdays(x) == "Saturday" | weekdays(x) == "Sunday")
  {y <- "Weekend"}
  else {y <- "Weekday"}
  y
})

``` 
 Make a panel plot containing a time series plot  of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
```{r}
 #Creating the data set that will be plotted using new weekday and weekend variables
activityByDay <-  aggregate(steps ~ interval + dayType, activity, mean, na.rm = TRUE)


library(ggplot2) 

#Plotting panel plot using ggplot2
dayPlot <-  ggplot(activityByDay, aes(x = interval , y = steps, color = dayType)) + 
  geom_line() + ggtitle("Average Daily Steps by Day Type") + 
  xlab("Interval") + 
  ylab("Average Number of Steps") +
  facet_wrap(~dayType, ncol = 1, nrow=2) +
  scale_color_discrete(name = "Day Type") 

print(dayPlot) 
 ``` 
